---
title: "gDRutils"
author: "gDR team"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{gDRutils}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(gDRutils)
suppressPackageStartupMessages(library(MultiAssayExperiment))
```

# Overview
`gDRutils` is the part of `gDR` suite. This package provides bunch of tools for, among others:
* data manipulation, especially output of the `gDRcore` package (`MultiAssayExperiments` and `SummarizedExperiment`),
* data extraction,
* managing identifiers used for creating `gDR` experiments,
* data validation.

# Use cases

## Data manipulation

The basic output of `gDRcore` package is the `MultiAssayExperiment` object. Function `MAEpply` allows for the data manipulation of this object, and can be used in a similar way as a basic function `lapply`.


```{r}
mae <- get_synthetic_data("finalMAE_combo_matrix_small.RDS")
MAEpply(mae, dim)
```
```{r}
MAEpply(mae, rowData)
```
This function allows also for extraction of unified data across all the `SummarizedExperiment`s inside `MultiAssayExperiment`, e.g.

```{r}
MAEpply(mae, rowData, unify = TRUE)
```

## Data extraction

All the metrics data are stored inside `assays` of `SummarizedExperiment`. For the downstream analyses we provide tools allowing for the extraction of the data into user-friendly `data.table` style.

There are two functions working on `MultiAssayExperiment` object (`convert_mae_assay_to_dt`) and for `SummarizedExperiment` (`convert_se_assay_to_dt`).

```{r}
mdt <- convert_mae_assay_to_dt(mae, "Metrics")
head(mdt, 3)
```
or alternatively for `SummarizedExperiment` object:

```{r}
se <- mae[[1]]
sdt <- convert_se_assay_to_dt(se, "Metrics")
head(sdt, 3)
```

## Managing gDR identifiers

In `gDR` we require standard identifiers that should be visible in the input data, such as e.g. `Gnumber`, `CLID`, `Concentration`. However, user can define their own custom identifiers.

To display gDR default identifier they can use `get_env_identifiers` function:

```{r}
get_env_identifiers()
```

To change any of these identifiers user can use `set_env_identifier`, e.g.

```{r}
set_env_identifier("concentration", "Dose")
```

and confirm, by displaying:

```{r}
get_env_identifiers("concentration")
```

To restore default identifiers user can use `reset_env_identifiers`.

```{r}
reset_env_identifiers()
```

```{r}
get_env_identifiers("concentration")
```

## Data validation

Applied custom changes in the gDR output can upset internal functions operation. Custom changes can be validated using `validate_MAE`

```{r}
validate_MAE(mae)
```

or `validate_SE`.

```{r}
validate_SE(se)
```

```{r, error=TRUE}
assay(se, "Normalized") <- NULL
validate_SE(se)
```



# SessionInfo {-}

```{r sessionInfo}
sessionInfo()
```
