---
title: "gDRutils"
author: "gDR team"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{gDRutils}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(gDRutils)
suppressPackageStartupMessages(library(MultiAssayExperiment))
```

# Overview
`gDRutils` is part of the `gDR` suite. This package provides a bunch of tools for, among others:

* data manipulation, especially output of the `gDRcore` package (`MultiAssayExperiments` and `SummarizedExperiment`),
* data extraction,
* managing identifiers used for creating `gDR` experiments,
* data validation.

# Use cases

## Data manipulation

The basic output of `gDRcore` package is the `MultiAssayExperiment` object. Function `MAEpply` allows for the data manipulation of this object, and can be used in a similar way as a basic function `lapply`.

```{r}
mae <- get_synthetic_data("finalMAE_combo_matrix_small")
MAEpply(mae, dim)
```
```{r}
MAEpply(mae, rowData)
```
This function allows also for extraction of unified data across all the `SummarizedExperiment`s inside `MultiAssayExperiment`, e.g.

```{r}
MAEpply(mae, rowData, unify = TRUE)
```

## Data extraction

All the metrics data are stored inside `assays` of `SummarizedExperiment`. For the downstream analyses we provide tools allowing for the extraction of the data into user-friendly `data.table` style.

There is a function working on the `MultiAssayExperiment` object as well as a set of functions working on the `SummarizedExperiment` object:

* convert_mae_assay_to_dt
* convert_se_assay_to_dt
* convert_se_assay_to_custom_dt
* convert_combo_data_to_dt


```{r}
mdt <- convert_mae_assay_to_dt(mae, "Metrics")
head(mdt, 3)
```
or alternatively for `SummarizedExperiment` object:

```{r}
se <- mae[[1]]
sdt <- convert_se_assay_to_dt(se, "Metrics")
head(sdt, 3)
```

## Managing gDR identifiers

### Overview

In `gDR` we require standard identifiers that should be visible in the input data, such as e.g. `Gnumber`, `CLID`, `Concentration`. However, user can define their own custom identifiers.

To display gDR default identifier they can use `get_env_identifiers` function:

```{r}
get_env_identifiers()
```

To change any of these identifiers user can use `set_env_identifier`, e.g.

```{r}
set_env_identifier("concentration", "Dose")
```

and confirm, by displaying:

```{r}
get_env_identifiers("concentration")
```

To restore default identifiers user can use `reset_env_identifiers`.

```{r}
reset_env_identifiers()
```

```{r}
get_env_identifiers("concentration")
```

### Validating identifiers

The `validate_identifiers` function checks if the specified identifier values exist in the data and (if needed) tries to modify them to pass validation.

```{r}
# Example data.table
dt <- data.table::data.table(
  Barcode = c("A1", "A2", "A3"),
  Duration = c(24, 48, 72),
  Template = c("T1", "T2", "T3"),
  clid = c("C1", "C2", "C3")
)

# Validate identifiers
validated_identifiers <- validate_identifiers(
  dt,
  req_ids = c("barcode", "duration", "template", "cellline")
)

print(validated_identifiers)
```

In detail, `validate_identifiers` wraps the following steps:

* modify identifier values to reflect the data, handling many-to-one mappings via the `.modify_polymapped_identifiers` function
* ensure that all required identifiers are present in the data via  the `.check_required_identifiers` function 
* check for polymapped identifiers in the data via the `.check_polymapped_identifiers` function
 
### Prettifying identifiers

Prettifying identifiers means making them more user-friendly and human-readable and is handled by the `prettify_flat_metrics` function. Please see [the relevant section](#prettifying) for more details.

```{r}
# Example of prettifying identifiers
x <- c("CellLineName", "Tissue", "Concentration_2")
prettified_names <- prettify_flat_metrics(x, human_readable = TRUE)
print(prettified_names)
```



## Data validation

Applied custom changes in the gDR output can disrupt internal functions operation. Custom changes can be validated using `validate_MAE`

```{r}
validate_MAE(mae)
```

or `validate_SE`.

```{r}
validate_SE(se)
```

```{r, error=TRUE, purl = FALSE}
assay(se, "Normalized") <- NULL
validate_SE(se)
```

There is also a group of functions to validate data used in the gDR application:

* is_combo_data
* has_single_codrug_data
* has_valid_codrug_data
* get_additional_variables

## Prettifying

Prettifying involves transforming data into a more descriptive and human-readable version. This is particularly useful for front-end applications where user-friendly names are preferred over technical or abbreviated terms.

In gdrplatform there are two entities that can be prettified:

* colnames of data.tables 
* assay names

### Colnames of data.table(s)

One can prettify the columns of the data.table(s) with a single function called `prettify_flat_metrics`.

```
dt <- get_testdata()[["raw_data"]]
colnames(dt)
prettify_flat_metrics(colnames(dt), human_readable = TRUE)
```

The `prettify_flat_metrics` function is in fact a wrapper for the following actions:

* conversion of the normalization-specific metric names via the `.convert_norm_specific_metrics` function
* moving the GDS source info to the end of the column name via the `.prettify_GDS_columns` 
* prettifying the metadata columns via the `.prettify_metadata_columns` function 
* prettifying the metric columns via the `.prettify_metric_columns` function 
* prettifying the co-treatment column names. via the `.prettify_cotreatment_columns` 
* minor corrections (removal of 'gDR' and  "_" prefixes, removal of spaces at the end/beginning, other)

In case of data.table(s) with combo excess and score assays some of the columns are prettified with the dedicated helper functions instead of using `prettify_flat_metrics`:

* get_combo_excess_field_names()
* get_combo_score_field_names() 

These helpers depend on the DATA_COMBO_INFO_TBL, (gDRutils) internal data.table.

### Assay names

The function `get_assay_names` is the primary solution for obtaining prettified versions of the assay names. It wraps the `get_env_assay_names` function which depends on ASSAY_INFO_TBL, (gDRutils) internal data.table.

There are some functions that wrap the `get_assay_names` function for combo data:

* get_combo_assay_names
* get_combo_score_assay_names
* get_combo_base_assay_names


# SessionInfo {-}

```{r sessionInfo}
sessionInfo()
```

